Linear and Differential Cryptanalysis of DES

### 2009-02-02 talk/draft1.tex
DifferentialCryptanalysis.java fixed ... Diff Slides modified ...

\subsection{Linear Attack: 4-round DES}
\begin{frame}
The plan for the DES Attacks:
\begin{itemize}
\item{Structurally and cryptographically, DES is similar to SPN.}
\item{We will adapt the linear attack on SPN to a linear attack on a 4-round DES algorithm and mount this attack.}
\item{We will also show how to extend this attack to a full 16-round DES algorithm and show it is faster than an exhaustive search.}
\end{itemize}
\end{frame}

\begin{frame}
The first step, as before, is to come up with a linear approximation for the S-Boxes used by DES.\\
\vspace{5mm}
Note that the S-Boxes here are 6-bit to 4-bit maps. We can use the same technique as before. For a given S-Box $a \in \{1, 2, \dots 8\}$, $1 \leq \alpha \leq 63$, $1 \leq \beta \leq 15$, we define $NS_{a}(\alpha, \beta)$ as the number of times (out of 64) for S-Box $a$ that for all the input patterns masked by $\alpha$ the output pattern masked by $\beta$ agrees with the value of S-Box $a$.
\[ NS_{a}(\alpha, \beta) = | \{ X | 0 \leq X < 64, ( \bigoplus_{i=0}^5 X[i]\cdot\alpha[i]) = (\bigoplus_{j=0}^3 S_a(X)[j] \cdot \beta[j])\} | \]
This looks more complicated, but is in fact the same expression as for SPN, except it takes into account all possible S-Boxes.
\end{frame}

\begin{frame}
Using $NS$ to find a linear approximation for the S-Boxes.\\
\vspace{5mm}
\begin{figure}
\includegraphics[totalheight=0.7\textheight]{NS_S5.PNG}
\caption{$NS_5(\alpha, \beta) - 32$}
\end{figure}
It turns out that the highest bias is for S-Box 5 with $\alpha = 16$ and $\beta = 15$
\end{frame}

\begin{frame}
From previous result, the S-Box approximation with highest bias for S-Box 5. Let $X$ be the input bits to the S-Box and let $Y$ be the output bits. Then:
\[ X_{2} = Y_{1} \oplus Y_{2} \oplus Y_{3} \oplus Y_{4} \]
Holds with bias of absolute value $\sim 0.31$\\
\vspace{5mm}
Tracing through the Feistel function to the right input bits $R$, round key bits $K$, and Feistel function output bits $F$ we get
\[ R_{17} \oplus F_{25} \oplus F_{14} \oplus F_{8} \oplus F_{3} = K_{26} \]

@@

\begin{frame}
Let $P_L$ and $P_R$ be the left and right plaintext bits. Let $C_L$ and $C_R$ be the left and right ciphertext bits. Let $K_i$ be the round key for round $i$ and $U_{4L}$ and $U_{4R}$ the left and right halves of the input to round 4.\\
\vspace{5mm}
Using the S-Box 5 approximation and the Piling Up Principle we can combine the approximations to get
\begin{align*}
 P_L[25] & \oplus P_L[14] \oplus P_L[8] \oplus P_L[3] \oplus \\
         & U_{4L}[25] \oplus U_{4L}[14] \oplus U_{4L}[8] \oplus U_{4L}[3] \oplus \\
			& P_R[17] \oplus U_{4R}[17] = K_1[26] \oplus K_3[26] \\
\end{align*}
The bias magnitude for this expression is
\[ (12/64)^2 + (1 - 12/64)^2 - 1/2 \approx 0.30 \]
This is the best expression for the 3-round DES.
\end{frame}

\begin{frame}
We perform the attack on the 4-round DES similarly to how we performed the attack on 4-round SPN, by ``undoing'' the last round.\\
\vspace{5mm}
\pause
Note that in the last round, $U_{4L}$ bits are not changed, only $U_{4R}$ bits are affected. Note also that there is only one $U_R$ bit the 3-round equation.
\begin{align*} 
P_L[25] \oplus P_L[14] \oplus P_L[8] \oplus & P_L[3] \oplus U_{4L}[25] \oplus U_{4L}[14] \oplus U_{4L}[8] \\ & \oplus U_{4L}[3] \oplus P_R[17] \oplus U_{4R}[17] = K_1[26] \oplus K_3[26]
\end{align*}
\vspace{5mm}
\pause
This bit, $U_{4R}[17]$ is affected by the result of S-Box 1 in round 4. If we guess correctly for the 6 partial subkey bits $K_{4,1}$ to $K_{4,6}$, the 3-round equation will hold with high bias, otherwise it will likely hold with close to 0 bias.\\
\vspace{5mm}
\pause
Thus, this attack will give us $6$ of the $56$ key bits. What this means in terms of time saved over exhaustive search is this: If it takes a year to find a key through exhaustive search of $2^{56}$ bits, after this attack it will only take $5$ days to exhaustively search the remaining $2^{50}$ bits.
\end{frame}

@@

\vspace{3mm}
Using the calculation from the previous slide, we see that for a 4-round DES, our equation holds with bias of about 1 in 4. Thus we will need about $16\times 4^2 = 256$ pairs for the attack to have near perfect chance of success.\\
\vspace{3mm}
For a full 16-round DES, the best approximation has bias $\sim 1.19\times 10^{-22}$, thus we would need on the order of $2^{47}$ plaintext-ciphertext pairs.
\end{frame}

\begin{frame}
The actual Attack on 4 round DES ... Demo ...
\end{frame}

\subsection{Differential Attack: 4-round DES}
\begin{frame}
Differential Attack Idea

\vspace{5mm}
Similar to the differential attack on SPN, here we care about the differential characteristics of the cipher. So, the basic idea is as follows

\begin{itemize}
\item{Construct the difference tables for the S-Boxes}
\item{Find the most probable difference pairs}
\item{Combine the pairs through all the rounds to get a highly probable differential characteristic until the penultimate round}
\item{Exploit this differential characteristic by choosing a number of pairs of plaintexts with the desired difference and undo the last round by guessing the keybits}
\item{If the initial choice of the input difference was smart, we will only need to guess a few keybits}
\item{The guess approving the differential characteristic with highest probability is correct}
\end{itemize}

@@

\begin{frame}
The chosen input difference:$\Delta P = 20 00 00 00 00 00 00 00$

Round 1 Differential

\[ \Delta P = 20 00 00 00 00 00 00 00 \rightarrow 20 00 00 00 00 00 00 00 = \Delta V_1 \]
Probability: $p_1 = 1$ as Feistel gives output difference $0$ for input $0$

Round 2 Differential

\[ \Delta U_2 = \Delta V_1 = 20 00 00 00 00 00 00 00 \rightarrow \]

this calculations and coding has not been completed yet ... will be done ...
\end{frame}

\begin{frame}
Differentials for SBoxes

@@

\begin{frame}
The actual Attack on 4 round DES ... Demo ...
\end{frame}

\subsection{Extension to full DES}
\begin{frame}
Extension of the same ideas to 6 / 8 / 16 round DES ... Probabilistic explanation ...

\section{Conclusion}
\begin{frame}
\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
Did we miss anything important?
\end{beamercolorbox}
\end{frame}

@@

\begin{frame}
What if that is false ? ... Discussion ...

\subsection{Links between the Attacks}
\begin{frame}
Links between Linear and Differential Attacks ...

\begin{frame}
Mathematically and while implementing ...

@@

\begin{frame}
If you are building a cryptosystem ... To Do / Not To Do ...

\begin{frame}
Questions ... Doubts ... Coding ... Misc ... Discussion ...

>>>>

\begin{frame}
\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
Linear Cryptanalysis of 4 Round DES
\end{beamercolorbox}
\end{frame}

\subsection{Linear Attack: 4-round DES}
\begin{frame}
The plan for the DES Attacks:
\begin{itemize}[<+->]
\item{Structurally and cryptographically, DES is similar to SPN.}
\item{We will adapt the linear attack on SPN to a linear attack on a 4-round DES algorithm and mount this attack.}
\item{We will also show how to extend this attack to a full 16-round DES algorithm and show it is faster than an exhaustive search.}
\end{itemize}
\end{frame}

\begin{frame}
The first step, as before, is to come up with a linear approximation for the S-Boxes used by DES.

\vspace{5mm}
\pause Note that the S-Boxes here are 6-bit to 4-bit maps. We can use the same technique as before.

\vspace{5mm}
\pause For a given S-Box $a \in \{1, 2, \dots 8\}$, $1 \leq \alpha \leq 63$, $1 \leq \beta \leq 15$, we define $NS_{a}(\alpha, \beta)$ as the number of times (out of 64) for S-Box $a$ that for all the input patterns masked by $\alpha$ the output pattern masked by $\beta$ agrees with the value of S-Box $a$.
\[ NS_{a}(\alpha, \beta) = | \{ X | 0 \leq X < 64, ( \bigoplus_{i=0}^5 X[i]\cdot\alpha[i]) = (\bigoplus_{j=0}^3 S_a(X)[j] \cdot \beta[j])\} | \]

\vspace{3mm}
\pause This looks more complicated, but is in fact the same expression as for SPN, except it takes into account all possible S-Boxes.
\end{frame}

\begin{frame}
Using $NS$ to find a linear approximation for the S-Boxes.
\vspace{2mm}
\begin{figure}
\includegraphics[totalheight=0.7\textheight]{NS_S5.PNG}
\caption{$NS_5(\alpha, \beta) - 32$}
\end{figure}
It turns out that the highest bias is for S-Box 5 with $\alpha = 16$ and $\beta = 15$
\end{frame}

\begin{frame}
So, we noticed the S-Box approximation with highest bias for S-Box 5.

\vspace{5mm}
\pause Letting $X$ be the input bits to S-Box 5 and $Y$ be the output bits,
\[ X_{2} = Y_{1} \oplus Y_{2} \oplus Y_{3} \oplus Y_{4} \]
holds with bias of absolute value $\sim 0.31$

\vspace{5mm}
\pause Tracing through the Feistel function to the right input bits $R$, round key bits $K$, and Feistel function output bits $F$ we get
\[ R_{17} \oplus F_{25} \oplus F_{14} \oplus F_{8} \oplus F_{3} = K_{26} \]

@@

\begin{frame}
Let $P_L$ and $P_R$ be the left and right plaintext bits. Let $C_L$ and $C_R$ be the left and right ciphertext bits. Let $K_i$ be the round key for round $i$ and $U_{4L}$ and $U_{4R}$ the left and right halves of the input to round 4.

\vspace{5mm}
\pause Using the S-Box 5 approximation and the Piling Up Principle we can combine the approximations to get
\begin{align*}
 P_L[25] & \oplus P_L[14] \oplus P_L[8] \oplus P_L[3] \oplus \\
         & U_{4R}[25] \oplus U_{4R}[14] \oplus U_{4R}[8] \oplus U_{4R}[3] \oplus \\
			& P_R[17] \oplus U_{4L}[17] = K_1[26] \oplus K_3[26] \\
\end{align*}

\pause The bias magnitude for this expression is
\[ (12/64)^2 + (1 - 12/64)^2 - 1/2 \approx 0.30 \]

\pause This is the best expression for the 3-round DES.
\end{frame}

\begin{frame}
\begin{columns}

\begin{column}{0.35\textwidth}
\begin{figure}
\includegraphics[totalheight=0.8\textheight]{DES_4.PNG}
\end{figure}
\end{column}

\begin{column}{0.65\textwidth}
\small
We perform the attack on the 4-round DES similarly to how we performed the attack on 4-round SPN, by ``undoing'' the last round.\\
\vspace{3mm}
\pause
Note that in the last round, $U_{4R}$ bits are not changed, only $U_{4L}$ bits are affected. Note also that there is only one $U_{4L}$ bit the 3-round equation.\\
\begin{align*} 
& P_L[25]    \oplus P_L[14]    \oplus P_L[8]    \oplus P_L[3] \oplus  \\
& U_{4R}[25] \oplus U_{4R}[14] \oplus U_{4R}[8] \oplus U_{4R}[3]  \oplus \\ 
& P_R[17]    \oplus U_{4L}[17] = K_1[26] \oplus K_3[26] \\
\end{align*}\\
\vspace{3mm}
\pause
This bit, $U_{4L}[17]$, is affected by the result of S-Box 1 in round 4. If we guess correctly for the 6 partial subkey bits $K_{4,1}$ to $K_{4,6}$, the 3-round equation will hold with high bias, otherwise it will likely hold with close to 0 bias.\\
\vspace{3mm}
\pause
Thus, this attack will give us $6$ of the $56$ key bits. If it takes a year to find a key through exhaustive search of $2^{56}$ bits, it only takes $5$ days to exhaustively search $2^{50}$ bits.
\end{column}

\end{columns}
\end{frame}

@@

\vspace{3mm}
Using the calculation from the previous slide, we see that for a 4-round DES, our equation holds with bias of about 1 in 4. \pause Thus we will need about $16\times 4^2 = 256$ pairs for the attack to have near perfect chance of success.

\vspace{3mm}
\pause For a full 16-round DES, the best approximation has bias $\sim 1.19\times 10^{-22}$, thus we would need on the order of $2^{47}$ plaintext-ciphertext pairs.
\end{frame}

\begin{frame}
DEMO: Attack on 4-round DES
\end{frame}

\subsection{Differential Attack: 4-round DES}
\begin{frame}
\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
Differential Cryptanalysis of 4 Round DES
\end{beamercolorbox}
\end{frame}

\begin{frame}
Differential Attack Idea

\vspace{5mm}
Similar to the differential attack on SPN, here we care about the differential characteristics of the cipher. \pause So, the basic idea is as follows

\begin{itemize}[<+->]
\item{Construct the difference tables for the S-Boxes}
\item{Find the most probable difference pairs}
\item{Combine the pairs through all the rounds to get a highly probable differential characteristic until the penultimate round}
\item{Exploit this differential characteristic by choosing a number of pairs of plaintexts with the desired difference and undo the last round by guessing the keybits}
\item{If the initial choice of the input difference was smart, we will only need to guess a few keybits}
\item{The guess approving the differential characteristic with highest probability is correct}
\end{itemize}

@@

\begin{frame}
\begin{columns}
\begin{column}{0.4\textwidth}
\begin{figure}
\includegraphics[totalheight=0.8\textheight]{des_diff_char.jpg}
\end{figure}
\end{column}

\begin{column}{0.6\textwidth}
Differential Characteristic of DES

\vspace{5mm}
Input Difference: $\Delta P = 20\:00\:00\:00\:00\:00\:00\:00$

\vspace{5mm}
\pause Propagating Difference Pairs
\begin{itemize}[<+->]
\item{Round 1: $S_i$ : $00 \rightarrow 00$}
\item{Round 2: $S_1$ : $04 \rightarrow 05$}
\item{Round 3: $S_4$ : $01 \rightarrow 05$ \\ $\qquad \qquad \: S_5$ : $10 \rightarrow 07$ \\ $\qquad \qquad \: S_8$ : $04 \rightarrow 07$}
\end{itemize}

\vspace{5mm}
\pause Differential Characteristic
\[ 20\:00\:00\:00\:00\:00\:00\:00 \rightarrow A0\:06\:18\:A0\:00\:00\:80\:02 \]

\end{column}
\end{columns}
\end{frame}

\begin{frame}
\begin{columns}
\begin{column}{0.4\textwidth}
\begin{figure}
\includegraphics[totalheight=0.8\textheight]{des_diff_char.jpg}
\end{figure}
\end{column}

\begin{column}{0.6\textwidth}
\vspace{2mm}
\pause Note: $00\:00\:80\:02$ implies difference of 2 bits which are both related to S-Box $S_1$ in round 4.

\vspace{2mm}
\pause Hence, in round 4
\[ D' = \Delta C_L \oplus 00\:00\:80\:02 \]
\pause and hence for $S_2, ... , S_8$, outputs are
\[ \Delta Y = P^{-1}(\Delta C_L) = S_{box}(E(C_R) \oplus K_4) \]

\vspace{2mm}
So, just guess the {\it target partial subkey} for $S_2, ... S_8$ individually, \pause and compute their predicted $\Delta Y$ with a number of plaintext pairs.

\vspace{2mm}
\pause This will ``undo'' round 4 and  the maximum probability of satisfying the 3 round differential characteristic will indicate a correct guess.
\end{column}
\end{columns}
\end{frame}

\begin{frame}
Probabilistic Analysis

\vspace{5mm}
Constructing the Difference Distribution tables for the S-Boxes and analyzing the probability distribution for differential propagation through SBoxes:

\vspace{2mm}
\pause
\begin{itemize}[<+->]
\item{Round 1: $S_i$ : $00 \rightarrow 00$ with $Pr = 1$}
\item{Round 2: $S_1$ : $04 \rightarrow 05$ with $Pr = 10/64$}
\item{Round 3: $S_4$ : $01 \rightarrow 05$  with $Pr = 16/64$ \\ $\qquad \qquad \: S_5$ : $10 \rightarrow 07$ with $Pr = 12/64$ \\ $\qquad \qquad \: S_8$ : $04 \rightarrow 07$ with $Pr = 12/64$}
\end{itemize}

\vspace{3mm}
\pause Differential Characteristic for $\Delta C_L$: $00\:00\:00\:00 \rightarrow 00\:00\:80\:02$ holds with probability
\[ p_D \: = \: 10/64 \: \approx \: 0.16 \: \gg 1/2^{32} \]

\vspace{2mm}
\pause
\begin{itemize}[<+->]
\item{How many pairs do we need?}
\item{Approximately $2^4 = 16$ for $100\%$ success}
\end{itemize}

@@

\begin{frame}
DEMO: Attack on 4-round DES
\end{frame}

\section{Conclusion}
\begin{frame}
\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
Concluding Remarks
\end{beamercolorbox}
\end{frame}

@@

\subsection{Links between the Attacks}
\begin{frame}
Linear Resistance:\\
\vspace{3mm}
Let $F: GF(2)^p \rightarrow GF(2)^q$.\\
\vspace{3mm}
Defn: A function F is bent if its Walsh transform $\mathcal{W}\{F\}$ is $\pm 2^{p/2}$.\\
\vspace{3mm}
Thm: A function F is perfect nonlinear (ie bias is 0) if and only if it is \textit{bent}. \\
\vspace{3mm}
Thm: Bent functions exist only for $p \geq 2q$ and $p$ even. \\
\vspace{5mm}
To sumarize: When $p\geq 2q$ and $p$ even, differential-resistant is equivalent to linear resistant, and to vectorial Bentness.
\end{frame}

@@

\begin{frame}
The problems of DES originated in two aspects:
\begin{itemize}
\item{6 bit S-Boxes are too easy to identify with linear or affine functions with high bias}
\item{Diffusion between rounds is just not fast enough}
\end{itemize}
AES, which replaced DES, provides an 8-Bit S-Box in the SubBytes step and higher diffusion through the ShiftRows and MixColumns steps. 
\end{frame}

\begin{frame}
Matsui rearrangement of S-Boxes ... Construction of S-Boxes ... IBM knew Diff 17 years earlier ... $10^44$ is closer to $10$ than to $10^45$ ...
\end{frame}

\begin{frame}
References:\\
\vspace{5mm}
M. Matsui, "Linear Cryptanalysis Method for DES Cipher", \textit{Advances in Cryptology - EUROCRYPT '93 (Lecture Notes in Computer Science no. 765),} Springer-Verlag, pp. 38\\
\vspace{2mm}
E Biham And A. Shamir, "Differential Cryptanalysis of DES-like Cryptosystems", \textit{Journal of Cryptology}, vol 4, no 1, pp. 3-72, 1991.\\
\vspace{2mm}
H. Heys, "A Tutorial on Linear and Differential Cryptanalysis"\\
\vspace{2mm}
F. Chabaud And S. Vaudenay, "Links Between Differential and Linear Cryptanalysis", \textit{Advances in Cryptology - EUROCRYPT '94 (Lexture Notes in Computer Science no. 950)}, Springer-Verlag, pp. 356-365, 1995.\\
\vspace{2mm}
J. Manferdelli, UW Cryptography Lecture Notes
\end{frame}

\begin{frame}
\begin{center}
The End.\\
Questions?
\end{center}
\end{frame}
